openTSNE: a modular \proglang{Python} library for t-SNE dimensionality reduction and embedding

### 2023-05-22 2023-05-revision-1/main.tex
Setup revision-1 folder

Evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations,
which makes it impractical with large data sets and
calls for efficient approximation schemes.
\citet{van2014accelerating} observed that the t-SNE gradient can be cast as an
N-body problem where data points represent particles that attract and repel each
other. The gradient from Equation~(\ref{eq:tsne_gradient}) can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
\Bigg ], \label{eq:grad_attr_rep} \notag
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
)^{-1}$. We can view this equation as a particle simulation, where the
two terms represent the attractive and repulsive forces between individual
particles. Each term lends itself to efficient approximations, enabling us to
reduce the time complexity of t-SNE to $\mathcal{O}(N)$. This allows modern
t-SNE implementations to be leveraged for the visualization of data sets
containing up to millions of data points.

@@

The t-SNE algorithm is non-parametric and does not define an explicit mapping from the
high-dimensional space to the embedding space. Embeddings of new data
points need to be found through optimization
schemes~\citep{policar2021embedding}. When adding new data points to an
existing, reference embedding, the reference data points are fixed in place
while new data points are allowed to find their respective positions. The
optimization remains the same as in standard t-SNE with only slight
modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \qquad
q_{j \mid i} = \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}, \notag
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Equation~(\ref{eq:symmetrize}). Plugging these terms into
Equation~(\ref{eq:kl_divergence}), we obtain the following gradient
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}. \notag
\label{eq:gradient}
\end{equation}

@@

Standard t-SNE reveals the clustering structure at a single level of resolution.
While the perplexity parameter can be used to control the trade-off between
local and global structure, this can be time-consuming, and small, well-defined
clusters can be missed. Alternatively, \citet{kobak2019heavy} suggest that
varying the degrees of freedom in the $t$-distribution can be used to explore
the clustering structure at different levels of resolution. Modifying
Equation~(\ref{eq:cauchy_kernel}) to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i -
\mathbf{y}_j ||^2 / \alpha \right )^{-\alpha}$ allows us to use heavier-tailed
distributions to model distances in the embedding space, which acts to highlight
small subgroups in the resulting visualization.

@@

We have used the following snippet to create Figs.~\ref{fig:cao}.a-c and to demonstrate how we can take advantage of the more advanced features of \opentsne to more quickly create meaningful visualizations of massive data sets. Here, we first take a sample of the data and use t-SNE to construct a globally consistent embedding, which we will then use for initialization of the positions for the remaining data points:
\begin{CodeChunk}
\begin{CodeInput}
# To improve the global structure of the final layout, we initialize the
# embedding by creating a t-SNE plot on a sample of points.

# Take a random sample of the data points.
indices = np.random.permutation(list(range(x.shape[0])))
reverse = np.argsort(indices)

x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]

# Create a t-SNE embedding on the sample, using a high perplexity value to
# emphasize the high level structure.
init_sample = openTSNE.TSNE(perplexity=500).fit(x_sample)

# We can next use the `transform` functionality to determine starting
# positions for the remaining data points. Since we don't actually want to
# optimize the remaining points separately from the sample points, we don't
# need to call the `transform` function. Instead, we'll call the
# `prepare_partial` function, which `transform` calls anyways under the hood.
init_rest = init_sample.prepare_partial(x_rest)

# Lastly, we merge `init_sample` and `init_rest` and restore the original
# ordering.
init_full = np.vstack((init_sample, init_rest))[reverse]
\end{CodeInput}
\end{CodeChunk}
We then generate the three different embeddings using the following code snippet (shortened for clarity):
\begin{CodeChunk}
\begin{CodeInput}
# We separately precompute the affinities as this can take quite a long time.
# This way, we can reuse the precomputed affinities in various different
# embeddings, without having to repeat the same, costly computation
# every time.
affinities = openTSNE.affinity.PerplexityBasedNN(x, perplexity=30)

# Create the TSNEEmbedding object using our previously calculated
# initialization and precomputed affinities.
embedding = openTSNE.TSNEEmbedding(init_full, affinities)

# Run the standard t-SNE early exaggeration phase
embedding_ee = embedding.optimize(n_iter=500, exaggeration=12, momentum=0.5)
embedding_c = embedding_ee.optimize(n_iter=500, exaggeration=4, momentum=0.8)
embedding_b = embedding_c.optimize(n_iter=500, exaggeration=2, momentum=0.8)
embedding_a = embedding_b.optimize(n_iter=500, exaggeration=1, momentum=0.8)
\end{CodeInput}
\end{CodeChunk}

@@

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a previously labeled single-cell data set and embed cells from a separate experiment into the reference landscape. The reference data from \citet{macosko2015highly} contains gene expression profiles from mouse retinal cells. By embedding the samples from a similar experiment on bipolar retinal cells by \citet{shekhar2016comprehensive}, we can correctly map the bipolar cell clusters onto the reference embedding. The following snippet shows the code necessary to generate Fig.~\ref{fig:macosko}:
\begin{CodeChunk}
\begin{CodeInput}
# We specify the old default parameter values which are often used to show
# why t-SNE misbehaves or fails to preserve global structure.
tsne_old = openTSNE.TSNE(
    initialization="random", learning_rate=200, n_iter=750
).fit(x)

# By default, openTSNE uses PCA-based initialization and automatically
# determines the best learning rate. This also allows us to lower the
# default number of iterations to 500, lowering the overall runtime
tsne_standard = openTSNE.TSNE(perplexity=[50, 500]).fit(x)

# Use the generated embeddings to add new data points
new_embedding = tsne_standard.transform(x_new)
\end{CodeInput}
\end{CodeChunk}

@@

\subsection{Speed\label{s:speed}}

>>>>

Evaluation of t-SNE gradients requires $\mathcal{O}(N^2)$ operations,
which makes it impractical with large data sets and
calls for efficient approximation schemes.
\citet{van2014accelerating} observed that the t-SNE gradient can be cast as an
N-body problem where data points represent particles that attract and repel each
other. The gradient from Equation~\ref{eq:tsne_gradient} can be rewritten as
\begin{equation}
\frac{\partial C}{\partial \mathbf{y}_i} = 4 \Bigg [
\underbrace{\sum_{j \neq i} p_{ij} q_{ij} Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{attractive forces}}  -
\underbrace{\sum_{j \neq i} q_{ij}^2 Z \left ( \mathbf{y}_i - \mathbf{y}_j \right )}_{\text{repulsive forces}}
\Bigg ], \label{eq:grad_attr_rep} \notag
\end{equation}
where $Z = \sum_{k \neq l}\left ( 1 + || \mathbf{y}_k - \mathbf{y}_l ||^2 \right
)^{-1}$. We can view this equation as a particle simulation, where the
two terms represent the attractive and repulsive forces between individual
particles. Each term lends itself to efficient approximations, enabling us to
reduce the time complexity of t-SNE to $\mathcal{O}(N)$. This allows modern
t-SNE implementations to be leveraged for the visualization of data sets
containing up to millions of data points.

@@

The t-SNE algorithm is non-parametric and does not define an explicit mapping from the
high-dimensional space to the embedding space. Embeddings of new data
points need to be found through optimization
schemes~\citep{policar2021embedding}. When adding new data points to an
existing, reference embedding, the reference data points are fixed in place
while new data points are allowed to find their respective positions. The
optimization remains the same as in standard t-SNE with only slight
modifications to $p_{ij}$ and $q_{ij}$
\begin{align}
p_{j \mid i} = \frac{\exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) /  \sigma_i^2 \right )}{\sum_{i} \exp \left ( -\frac{1}{2} \mathcal{D}(\mathbf{x}_i, \mathbf{v}_j) / \sigma_i^2 \right )}, \qquad
q_{j \mid i} = \frac{\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}{\sum_{i}\left ( 1 + || \mathbf{y}_i - \mathbf{w}_j ||^2 \right )^{-1}}, \notag
\end{align}
\noindent where $\mathbf{V} = \left \{ \mathbf{v}_1, \mathbf{v}_2, \dots,
\mathbf{v}_M \right \} \in \mathbb{R}^D$ where $M$ is the number of samples in
the new data set and $\mathbf{W} = \left \{ \mathbf{w}_1, \mathbf{w}_2, \dots,
\mathbf{w}_M \right \} \in \mathbb{R}^d$. Additionally, we omit the
symmetrization step in Equation~\ref{eq:symmetrize}. Plugging these terms into
Equation~\ref{eq:kl_divergence}, we obtain the following gradient
\begin{equation}
\frac{\partial C}{\partial \mathbf{w}_j} = 2 \sum_i \left ( p_{j \mid i} - q_{j \mid i} \right ) \left ( \mathbf{y}_i - \mathbf{w}_j \right ) \left ( 1 + || \mathbf{y}_i - \mathbf{w}_j || ^2 \right )^{-1}. \notag
\label{eq:gradient}
\end{equation}

@@

Standard t-SNE reveals the clustering structure at a single level of resolution.
While the perplexity parameter can be used to control the trade-off between
local and global structure, this can be time-consuming, and small, well-defined
clusters can be missed. Alternatively, \citet{kobak2019heavy} suggest that
varying the degrees of freedom in the $t$-distribution can be used to explore
the clustering structure at different levels of resolution. Modifying
Equation~\ref{eq:cauchy_kernel} to $q_{ij} \propto \left ( 1 + || \mathbf{y}_i -
\mathbf{y}_j ||^2 / \alpha \right )^{-\alpha}$ allows us to use heavier-tailed
distributions to model distances in the embedding space, which acts to highlight
small subgroups in the resulting visualization.

@@

We have used the following code to create Figs.~\ref{fig:cao}.a-c and to demonstrate how we can take advantage of the more advanced features of \opentsne to more quickly create meaningful visualizations of massive data sets.

First, we take small subset of data points and create a t-SNE embedding using a high perplexity value to emphasize the high level structure:
\begin{CodeChunk}
\begin{CodeInput}
indices = np.random.permutation(list(range(x.shape[0])))
x_sample, x_rest = x[indices[:25000]], x[indices[25000:]]

init_sample = openTSNE.TSNE(perplexity=500).fit(x_sample)
\end{CodeInput}
\end{CodeChunk}
Next, we will use this embedding to determine the starting positions for the remaining data points:
\begin{CodeChunk}
\begin{CodeInput}
init_rest = init_sample.prepare_partial(x_rest)
\end{CodeInput}
\end{CodeChunk}
Lastly, we merge \code{init_sample} and \code{init_rest} and restore the original ordering to obtain the full initialization:
\begin{CodeChunk}
\begin{CodeInput}
init_full = np.vstack((init_sample, init_rest))[np.argsort(indices)]
\end{CodeInput}
\end{CodeChunk}

Next, we precompute the affinity model. This can take quite a long time and, this way, we can reuse the same affinity model in various different embeddings, without having to repeat the same, costly computation every time:
\begin{CodeChunk}
\begin{CodeInput}
affinities = openTSNE.affinity.PerplexityBasedNN(x, perplexity=30)
\end{CodeInput}
\end{CodeChunk}

Using our precomputed initialization and affinity model, we are now ready to create our \code{TSNEEmbedding} object and optimize the embedding to our liking: 
\begin{CodeChunk}
\begin{CodeInput}
embedding = openTSNE.TSNEEmbedding(init_full, affinities)

embedding_ee = embedding.optimize(n_iter=500, exaggeration=12, momentum=0.5)
embedding_c = embedding_ee.optimize(n_iter=500, exaggeration=4, momentum=0.8)
embedding_b = embedding_c.optimize(n_iter=500, exaggeration=2, momentum=0.8)
embedding_a = embedding_b.optimize(n_iter=500, exaggeration=1, momentum=0.8)
\end{CodeInput}
\end{CodeChunk}

@@

Figs.~\ref{fig:macosko}.b and \ref{fig:macosko}.c demonstrate how we can use a previously labeled single-cell data set and embed cells from a separate experiment into the reference landscape. The reference data from \citet{macosko2015highly} contains gene expression profiles from mouse retinal cells. By embedding the samples from a similar experiment on bipolar retinal cells by \citet{shekhar2016comprehensive}, we can correctly map the bipolar cell clusters onto the reference embedding.

The following snippet shows the code necessary to generate Fig.~\ref{fig:macosko}.

We first construct Fig.~\ref{fig:macosko}.a with the most commonly used parameter values which aims to show that these can lead to poor cluster separation and an overall lack of global consistency:
\begin{CodeChunk}
\begin{CodeInput}
tsne_old = openTSNE.TSNE(
    initialization="random", learning_rate=200, n_iter=750
).fit(x)
\end{CodeInput}
\end{CodeChunk}
We generate the embedding from Fig.~\ref{fig:macosko}.b using parameter values emphasizing the global coherence of the resulting visualization. By default, \opentsne uses PCA-based initialization and automatically determines the optimal learning rate. This allows \opentsne to lower the default number of iterations to 500. To emphasize the global consistency of the resulting embedding, we can also specify multiple perplexity values, resulting in a multiscale affinity model:
\begin{CodeChunk}
\begin{CodeInput}
tsne_multiscale = openTSNE.TSNE(perplexity=[50, 500]).fit(x)
\end{CodeInput}
\end{CodeChunk}

Finally, we can embed new samples into our obtained embedding space:
\begin{CodeChunk}
\begin{CodeInput}
new_embedding = tsne_standard.transform(x_new)
\end{CodeInput}
\end{CodeChunk}

@@

\subsection{Speed}\label{s:speed}
